Formula translation in C++, Python and modern Fortran: a~case~study~of~the~language~choice tradeoffs

### 2013-07-01 paper/doc/paper.tex
Revert "hint on the CMake 2.8.10 bug"
This reverts commit d0d49d3acb5ecb007858dad7d618d065668bf578.
initial changes before planned submission to GMD

%%%%% <cpc>
\documentclass[review,peprint,10pt,5p,times]{elsarticle}
\usepackage[switch]{lineno}
\addtolength{\hoffset}{.5cm}
%%%%% </cpc>

@@

@@

\newcommand{\prog}[1]{{\rm\bf#1}}
\newcommand{\url}[1]{{#1}}

@@

\begin{document}
  \begin{frontmatter}

    \title{
      Object-oriented implementations of the MPDATA advection equation solver in~C++,~Python~and~Fortran
    }

@@

    \address[1]{Institute of Geophysics, Faculty of Physics, University of Warsaw}
    \address[2]{PyPy Team}

    \begin{abstract}
        Three object-oriented implementations of a prototype solver of the advection equation are introduced.
        The presented programs are based on Blitz++ (C++), NumPy (Python), and Fortran's built-in array containers.
        The solvers include an implementation of the Multidimensional Positive-Definite 
          Advective Transport Algorithm (MPDATA).
        The introduced codes exemplify how the application of 
          object-oriented programming (OOP) techniques allows to reproduce the mathematical notation 
          used in the literature within the program code.
        A discussion on the tradeoffs of the programming language choice is presented.
        The main angles of comparison are code brevity and syntax clarity
          (and hence maintainability and auditability) as well as performance.
        In the case of Python, a significant performance gain is observed when switching from the standard 
          interpreter (CPython) to the PyPy implementation of Python.
        Entire source code of all three implementations is embedded in the text and is licensed
          under the terms of the GNU GPL license. 
    \end{abstract}

    % up to six in Comp. Phys. Comm.
    \begin{keyword}
      object-oriented programming, advection equation, MPDATA, C++, Fortran, Python
    \end{keyword}

  \end{frontmatter}

%%%%% <cpc>
\linenumbers % for the review
\section*{Program summary}
\noindent
{\it Program title:} mpdata-oop\\
{\it Licensing provisions:} GNU GPL version 3\\
{\it Programming languages:} C++, Python and Fortran\\
{\it Computer:} hardware-independent (tested on x86\_64)\\
{\it Operating system:} OS-independent (developed on GNU/Linux)\\
{\it RAM/CPU time requirements:} problem-dependent\\
{\it Has the code been vectorised or parallelised?:}\\ 
  GCC 4.8 auto-vectorises parts of the Fortran and the C++ code.\\
  The code is readily adaptable for concurrent computation using
  domain decomposition.\\
{\it External libraries used:} Blitz++, NumPy\\
{\it Classification:} 70 (Mathematical and Numerical Methods)\\
{\it Nature of problem:} advective transport\\
{\it Solution method:} Multidimensional Positive Definite Advective Transport Algorithm (MPDATA)\\
{\it Reference:} Smolarkiewicz 1984, J. Comp. Phys. 54, 325--362
%%%%% </cpc>

  \tableofcontents

  \section{Introduction}

@@

  Application of OOP techniques in development of numerical modelling software may help to:
  \begin{enumerate}[label=(\roman*), leftmargin=*, widest=ii]
    \item{maintain modularity and separation of program logic layers (e.g. separation of
      numerical algorithms, parallelisation mechanisms, data input/output, error handling and
      the description of physical processes); and}
    \item{{\bf shorten and simplify the source code and improve its readability by reproducing within 
      the program logic the mathematical notation used in the literature}.}
  \end{enumerate}
  The first application is attainable, yet arguably cumbersome, with procedural programming.
  The latter, virtually impossible to obtain with procedural programming, is the focus of this paper.
  It also enables the compiler or library authors to relieve the user (i.e. scientific programmer)
    from hand-coding optimisations, a practice long recognised as having {\em a strong negative impact when debugging
    and maintenance are considered} \citep{Knuth_1974}.

  MPDATA \citep{Smolarkiewicz_1984} stands for Multidimensional Positive Definite Advective Transport Algorithm and is
    an example of a numerical procedure used in weather, climate and ocean simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002}.
  MPDATA is a solver for systems of advection equations of the following form:
  \begin{equation}\label{eq:adv}
    \input{formulae_adv}
  \end{equation}
  that describe evolution of a scalar field $\psi$ transported by the fluid flow with velocity $\vec{v}$.
  Quoting Numerical Recipes once more, development of methods to numerically solve such problems 
    {\em ''is an art as much as a science''} \citep[][Sec.~20.1]{Press_et_al_2007},
    and MPDATA is an example of the state-of-the art in this field.
  MPDATA is designed to accurately solve equation (\ref{eq:adv}) in an arbitrary
    number of dimensions assuring positive-definiteness of scalar field $\psi$ 
    and incurring small numerical diffusion.
  All relevant MPDATA formul\ae~are given in the text but are presented without
    derivation or detailed discussion.
  For a recent review of MPDATA-based techniques see \citet[][and references therein]{Smolarkiewicz_2006}.

  In this paper we introduce and discuss object-oriented implementations of an MPDATA-based 
    two-dimensional (2D) advection equation solver written in C++11 (\citeauthor{ISO_14882} 14882:2011), 
    Python \citep{Rossum_and_Drake_2011} and Fortran 2008 (\citeauthor{ISO_1539-1} 1539-1:2010). 
  In the following section we introduce the three implementations
    briefly describing the algorithm itself and
    discussing where and how the OOP techniques may be applied in its implementation.
  The syntax and nomenclature of OOP techniques are used without introduction,
    for an overview of OOP in context of C++, Python and Fortran, consult for example
    \citep[][Part~II]{Stroustrup_2000}, \citep[][Chapter~5]{Pilgrim_2004} and
    \citep[][Chapter~11]{Markus_2012}, respectively.
  The third section of this paper covers performance evaluation of the three implementations.
  The fourth section covers discussion of the tradeoffs of the programming language choice.
  The fifth section closes the article with a brief summary.

  Throughout the paper we present the three implementations by discussing 
    source code listings which cover the entire program code.
  Subsections \ref{sec:array}-\ref{sec:solver} describe all three implementations,
    while subsequent sections \ref{sec:cyclic}-\ref{sec:example} cover discussion of C++ code 
    only.
  The relevant parts of Python and Fortran codes do not differ significantly, and for readability reasons 
    are presented in \ref{app:P} and \ref{app:F}, respectively.

@@

  All codes are structured in a way allowing compilation of the code 
    in exactly the same order as presented in the text within one source file,
    hence every Fortran listing contains definition of a separate module.

@@

  Solution of equation (\ref{eq:adv}) using MPDATA implies discretisation onto a grid of the $\psi$ and 
    the Courant number $\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}$ fields, where $\Delta t$ is the 
    solver timestep and $\Delta x$ is the grid spacing.

@@

  The authors do expect some performance gain could 
    be obtained by introducing into the codes some ''manual'' optimisations -- 
    code rearrangements aimed solely at the purpose of increasing performance.
  These were avoided intentionally as they degrade code readability,
    should in principle be handled by the compilers,
    and are generally advised to be avoided \citep[e.g.][section 3.12]{bib_CERNcpp}.

@@

  It was shown in section~\ref{sec:impl} that C++11/Blitz++, Python/NumPy and Fortran 2008
    provide comparable functionalities in terms of matching the blackboard abstractions
    within the program code.
  Taking into account solely the part of code representing particular formul\ae~
    (e.g. listings C.21, P.17, F.20 and equation \ref{eq:antidiff1}) all three
    languages allow to match (or surpass) \LaTeX~in its brevity of formula translation syntax.
  All three languages were shown to be capable of providing mechanisms to compactly represent such abstractions as:
  \begin{itemize}
    \item{loop-free array arithmetics;}
    \item{definitions of functions returning array-valued expressions;}
    \item{permutations of array indices allowing dimension-independent definitions
      of functions (see e.g. listings C.12 and C.13, P.10 and P.11, F.11 and F.12);}
    \item{fractional indexing of arrays corresponding to employment of a staggered grid.}
  \end{itemize}
  Three issues specific to Fortran that 
    resulted in employment of a more repetitive or cumbersome syntax than in C++ or Python
    were observed:
  \begin{itemize}
    \item{Fortran does not feature a mechanism allowing to reuse a single piece of code (algorithm)
      with different data types (compare e.g. listings C.6, P.5 and F.4) such as
      templates in C++ and the so-called duck typing in Python;}
    \item{Fortran does not allow function calls to appear on the left hand side
      of assignment (see e.g. how the \prog{ptr} pointers were used as a workaround in the \prog{cyclic\_fill\_halos}
      method in listing F.8);}
    \item{Fortran lacks support for arrays of arrays (cf. Sect. \ref{sec:sequence}).}
  \end{itemize}
  Interestingly, the limitation in extendability via inheritance was found to
    exist partially in NumPy as well (see footnote \ref{fnt:slice}).
  The lack of a counterpart in Fortran to the C++ template mechanism was identified in
    \citep{Cary_et_al_1997}
    as one of the key deficiencies of Fortran when compared with C++ in context 
    of applicability to object-oriented scientific programming.

@@

  \section{Summary and outlook}

@@

  \section*{Acknowledgements}
  {\small
    \noindent 
    We thank Piotr Smolarkiewicz and Hanna Pawłowska for their help throughout the project.
    This study was partly inspired by the lectures of Lech Łobocki.

    \noindent
    Development of NumPy support in PyPy was led by Alex Gaynor, Matti Picus and MF.
  }

@@

  \setcounter{section}{15} 
  \section{Python code for sections \ref{sec:cyclic}--\ref{sec:mpdatasolver}}\label{app:P}

@@

  \setcounter{section}{5} 
  \section{Fortran code for sections \ref{sec:cyclic}--\ref{sec:mpdatasolver}}\label{app:F}

@@

  \bibliographystyle{model1-num-names}
  \bibliography{paper}

>>>>

@@

%%%%% <gmd>
\documentclass[gmd]{copernicus}%_discussions}
%%%%% </gmd>

@@

\newcommand{\prog}[1]{{\rm\bf#1}}

@@

\begin{document}
    \linenumbers

    \title{
      Formula translation in C++, Python and modern Fortran: a~case~study~of~the~language~choice tradeoffs
    }

@@

    \affil[1]{Institute of Geophysics, Faculty of Physics, University of Warsaw}
    \affil[2]{PyPy Team}

    \correspondence{Sylwester Arabas (sarabas@igf.fuw.edu.pl)}
    \runningauthor{Arabas, Jarecka et al.}
    \runningtitle{Formula translation in C++, Python \& Fortran}

    \maketitle

    \begin{abstract}
        Three object-oriented implementations of a prototype solver of the advection equation are introduced.
        The presented programs are based on Blitz++ (C++), NumPy (Python), and Fortran's built-in array containers.
        The solvers constitute implementations of the Multidimensional Positive-Definite 
          Advective Transport Algorithm (MPDATA).
        The introduced codes serve as exemples for how the application of 
          object-oriented programming (OOP) techniques allows to reproduce the mathematical notation 
          used in the literature within the program code.
        A discussion on the tradeoffs of the programming language choice is presented.
        The main angles of comparison are code brevity and syntax clarity
          (and hence maintainability and auditability) as well as performance.
        All performance tests are performed using free and open-source compilers.
        In the case of Python, a significant performance gain is observed when switching from the standard 
          interpreter (CPython) to the PyPy implementation of Python.
        Entire source code of all three implementations is embedded in the text and is licensed
          under the terms of the GNU GPL license. 
    \end{abstract}

%    \begin{keyword}
%      object-oriented programming, advection equation, MPDATA, C++, Fortran, Python
%    \end{keyword}

  \introduction

@@

  Application of OOP techniques in development of numerical modelling software may help to:
  \begin{enumerate}[label=(\roman*), leftmargin=*, widest=ii]
    \item{maintain modularity and separation of program logic layers (e.g. separation of
      numerical algorithms, parallelisation mechanisms, data input/output, error handling and
      the description of physical processes); and}
    \item{{\bf shorten and simplify the source code and improve its readability by reproducing within 
      the program logic the mathematical notation used in the literature}.}
  \end{enumerate}
  The first application is attainable, yet arguably cumbersome, with procedural programming.
  The latter, virtually impossible to obtain with procedural programming, is the focus of this paper.

  The aim of this paper is to show how OOP techniques can be used to represent mathematical
    notation within the program code.
  For discussion purposes, three sample implementations of a numerical scheme for solving
    the advection equation are introduced.
  The three programs are written in C++, Python and Fortran. 
  They implement a two-dimensional version of the 
    Multidimensional Positive Definite Advective Transport Algorithm \citep[MPDATA][]{Smolarkiewicz_1984}.
  MPDATA is an example of a numerical procedure used in weather, climate and ocean simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002}.
  Even in its most basic form presented herein, MPDATA is complex enough to contain
    a wide range of mathematical abstractions that can be represented using OOP constructs, 
    yet it is simple enough to allow inclusion of the entire source code within the paper text.
  All relevant MPDATA formul\ae~are given in the text alongside corresponding code fragments.
  In compliance with the article focus, these formul\ae~are presented without derivation or detailed discussion.
  For a recent review of MPDATA-based techniques including an introductory description of the algorithm
    and a thorough list of references, see \citet[][]{Smolarkiewicz_2006}.

  In the following section we introduce the three implementations
    briefly describing the algorithm itself and
    discussing where and how the OOP techniques may be applied in its implementation.
  Subsections \ref{sec:array}-\ref{sec:solver} describe all three implementations,
    while subsequent sections \ref{sec:cyclic}-\ref{sec:example} cover discussion of C++ code only.
  The relevant parts of Python and Fortran codes do not differ significantly, and for readability reasons 
    are presented in \ref{app:P} and \ref{app:F}, respectively.
  The third section of this paper covers performance evaluation of the three implementations.
  The fourth section covers discussion of the tradeoffs of the programming language choice.
  The fifth section closes the article with a brief summary.

@@

  All codes are structured in a way allowing compilation of the code 
    in exactly the same order as presented in the text within one source file,
    hence every Fortran listing contains definition of a separate module.
  The language syntax and OOP nomenclature are used without introduction in the paper,
    for an overview of OOP in context of C++, Python and Fortran, consult for example
    \citep[][Part~II]{Stroustrup_2000}, \citep[][Chapter~5]{Pilgrim_2004} and
    \citep[][Chapter~11]{Markus_2012}, respectively.

@@

  MPDATA is a solver for systems of advection equations of the following form:
  \begin{equation}\label{eq:adv}
    \input{formulae_adv}
  \end{equation}
  that describe evolution of a scalar field $\psi$ transported by the fluid flow with velocity $\vec{v}$.
  Solution of equation (\ref{eq:adv}) using MPDATA implies discretisation onto a grid of the $\psi$ and 
    the Courant number $\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}$ fields, where $\Delta t$ is the 
    solver timestep and $\Delta x$ is the grid spacing.

@@

@@

  It was shown in section~\ref{sec:impl} that C++11/Blitz++, Python/NumPy and Fortran 2008
    provide comparable functionalities in terms of matching the blackboard abstractions
    within the program code.
  Taking into account solely the part of code representing particular formul\ae~
    (e.g. listings C.21, P.17, F.20 and equation \ref{eq:antidiff1}) all three
    languages allow to match (or surpass) \LaTeX~in its brevity of formula translation syntax.
  All three languages were shown to be capable of providing mechanisms to compactly represent such abstractions as:
  \begin{itemize}
    \item{loop-free array arithmetics;}
    \item{definitions of functions returning array-valued expressions;}
    \item{permutations of array indices allowing dimension-independent definitions
      of functions (see e.g. listings C.12 and C.13, P.10 and P.11, F.11 and F.12);}
    \item{fractional indexing of arrays corresponding to employment of a staggered grid.}
  \end{itemize}

  Making use of features such as loop-free arithmetics not only shortens the code,
    but also enables the compiler or library authors to relieve the user (i.e. scientific programmer)
    from hand-coding optimisations (e.g. loop order choice).
  Hand-coded optimisations -- code rearrangements aimed solely at the purpose of increasing performance --
    were long recognised as having {\em a strong negative impact when debugging
    and maintenance are considered} \citep{Knuth_1974},
    and are generally advised to be avoided \citep[e.g.][section 3.12]{bib_CERNcpp}.

  Three issues specific to Fortran that 
    resulted in employment of a more repetitive or cumbersome syntax than in C++ or Python
    were observed:
  \begin{itemize}
    \item{Fortran does not feature a mechanism allowing to reuse a single piece of code (algorithm)
      with different data types (compare e.g. listings C.6, P.5 and F.4) such as
      templates in C++ and the so-called duck typing in Python;}
    \item{Fortran does not allow function calls to appear on the left hand side
      of assignment (see e.g. how the \prog{ptr} pointers were used as a workaround in the \prog{cyclic\_fill\_halos}
      method in listing F.8);}
    \item{Fortran lacks support for arrays of arrays (cf. Sect. \ref{sec:sequence}).}
  \end{itemize}
  Interestingly, the limitation in extendability via inheritance was found to
    exist partially in NumPy as well (see footnote \ref{fnt:slice}).
  The lack of a counterpart in Fortran to the C++ template mechanism was identified in
    \citep{Cary_et_al_1997}
    as one of the key deficiencies of Fortran when compared with C++ in context 
    of applicability to object-oriented scientific programming.

@@

  \conclusions[Summary and outlook]

@@

@@

  %\setcounter{section}{15} 
  \section{Python code for sections \ref{sec:cyclic}--\ref{sec:mpdatasolver}}\label{app:P}

@@

  %\setcounter{section}{5} 
  \section{Fortran code for sections \ref{sec:cyclic}--\ref{sec:mpdatasolver}}\label{app:F}

@@

  \begin{acknowledgements}
    \noindent 
    We thank Piotr Smolarkiewicz and Hanna Pawłowska for their help throughout the project.
    This study was partly inspired by the lectures of Lech Łobocki.

    \noindent
    Development of NumPy support in PyPy was led by Alex Gaynor, Matti Picus and MF.
  \end{acknowledgements}

  \bibliographystyle{copernicus}
  \bibliography{paper}
